Supplementary Information: Drought sensitivity is not a species trait: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-10-10 doc/response_to_reviewer_R1.Rmd

---
title: 'Responses to the Reviewer'
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
suppress-bibliography: TRUE
---

Dear Editor,

We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" (ELE-00826-2025) for consideration for publication in Ecology Letters. 
We have addressed all points raised by the reviewers as detailed in our responses to the reviewers. 
The most significant changes were discussing global implications as well as species effects in more detail. 
We have also revised the methodology and results to highlight the correspondence between the predictions and the results. 
We have also updated Figures 1 and 2 as well as the relevant figures in the supplementary material as requested by the reviewers.
Finally, following a friendly review by a collaborator, we also present an alternate, analogous set of causal inquiry models in the supplementary information that supports the conclusions in the main text.

@@

•	It is always helpful to know the study site. I recommend including some pictures of the forest and study area.

**RESPONSE: Thanks for this comment. This is a great idea, and we would like to include this as well to provide further context to readers. Given that we are at the limit of display items in the main text, we have included this as Figure S1 in the supplementary material.**

•	The methodology should clearly state whether the 30 species selected for the study represent the functional groups. It should include the number of species per Deciduousness group.

**RESPONSE: Thanks for this comment. We agree that it is useful to include the functional representativeness of the species. As we mention in the methods, our metric of deciduousness is a continuous variable and not categorical. And so, reporting the number per deciduousness group would require arbitrarily discretising this continuous variable. Instead, we now report a supplementary figure (Figure S9) showing the distribution of deciduousness among the selected species.**

@@

**RESPONSE: Thanks for this suggestion. We were excited to see this global study published recently. We have now discussed our focused work in relation to this global study based on the alignment of the key message - the diversity of drought responses with drought types and species traits (deciduousness). However, dendroband datasets can be key to filling gaps in spatial and climatic coverage of tree rings records in the tropics (reviewed by @groenendijk_importance_2025). Moreover, our results showcasing response diversity among individuals within a species demonstrates an important caveat to aggregate/species-chronology-based inferences from tree-ring records.**

•	There was something missing to discuss about the species in particular, some that have been more affected than others and why, especially those in conservation categories.

**RESPONSE: Thanks for this suggestion. We have now included a supplementary table (Table S1) that details the response of specific species to specific drought events. This summary highlights our main message that the species responses are dissimilar across drought events. This table also includes available information on the IUCN categories of the species analysed (accessed in September 2025) where information was avaialable. Two of the species analysed are "Endangered", including *Afzelia xylocarpa*, the species that was discussed. There are also "Vulnerable" and "Data Deficient" species in the list, some of which have consistent negative drought sensitivities. This table would therefore be useful for forest managers and conservation professionals and to guide species-specific research directions. We have also added a line highlighting the contrasts between species responses in the discussion, which now reads:**

**L421-430: "An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig. S5, Table S1).
Extrapolating from such dendrochronological data, therefore, could lead to important errors if attributed to other species, potentially misrepresenting drought impacts conservation or silviculture, e.g. *Dipterocarpus alatus*, a conservation and commercial priority species, presents a different drought response (Table S1)."**

@@

**As to why different species might be responding differently, a key difference could be their evolutionary strategies along hydraulic safety vs. efficiency. 
We have discussed our results in light of deciduousness as an example.
However, we recognise that many other above and belowground traits may contribute to these differences and need to examined further.
We have edited text in the discussion to read:**

**L400-404: "Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."**

@@

![](display/Fig2_alt2.png)

339-346: I had a hard time understanding this result, on the one hand it says that Crown exposure was associated with greater growth declines in 2010, but not for 2015 and 2020. However, according Fig 5b all all crown exposure categories had negative sensitivities in 2015 and 2020.

**RESPONSE: We agree the wording could be clearer. Our focus is on how growth sensitivity changes with crown exposure. Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity. In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees. We have now clarified this in the text which now reads:**

**L340: "Predicted effects of CII on sensitivity decreased monotonically, with strong decreases in 2010, weak decrease in 2015, and no effect in 2020."**

@@

Fig. S1 What does the shade area indicate? The three different drought events?

**RESPONSE: Yes, the shaded area indicates the three different drought years. We have edited the text to reflect this.**

Fig S6 The R Square is too low, so it is difficult to detect the linear relationship. The same issue for Fig. S11 and Fig. S12. For the non-significant correlation, it’s better use dashed lines instead of solid line.

**RESPONSE: Thanks for this suggestion. The dotted line in Fig S6 was originally the 1:1 line - a strict null expectation. We have now added dashed lines for the significant correlations. Following the same logic, for S11 and S12, we have now removed lines for the non-significant correlations, but added Friedman's SuperSmoother fits for all plots to visually demonstrate the direction of the relationship.**

### 2025-10-10 doc/supplementary_information.Rmd
cosmetic changes

Map of trees

@@

![**Figure S9:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/twi.png)

![**Figure S10:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/dbh_cii.png)

@@

![**Figure S11:** Distribution of raw sensitivity values across years and their relationship with DBH, CII and TWI. Blue lines are linear model fits for statistically significant relationships at p<0.05. Red lines are Friedman's SuperSmoother fits for all.](display/sens_var_rels.png)

![**Figure S12:** Distribution of residuals of species intercept models of sensitivity and their relationship with scaled DBH, TWI and CII and deciduousness. Grey lines are linear model fits for statistically significant relationships at p<0.05. Species intercept model results are reported in main text.](display/residuals.png)

@@

![**Figure S13:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)

@@

![**Figure S14:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/cond_dep_dbh_twi.png)

![**Figure S15:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/cond_dep_cii_twi.png)

@@

![**Figure S16:** Correlation of modelled species sensitivities with Topographic Position Index](display/FigSI_TPI_preds.png)

@@

![**Figure S17:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TPI.](display/Fig4_tpi_3yrs.png)

@@

![**Figure S18:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)

@@

![**Figure S19:** Model results from three independent causal inquiry models of sensitivity following backdoor criteria for total causal effect of DBH, CII and TWI with species random effect on intercept and slope.](display/causal_inquiry.png)

>>>>

@@

![**Figure S9:** Distribution of deciduousness values among the 30 species analysed](display/Fig_dec_dist.png)

![**Figure S10:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/twi.png)

![**Figure S11:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/dbh_cii.png)

@@

![**Figure S12:** Distribution of raw sensitivity values across years and their relationship with DBH, CII and TWI. Blue lines are linear model fits for statistically significant relationships at p<0.05. Red lines are Friedman's SuperSmoother fits for all.](display/sens_var_rels.png)

![**Figure S13:** Distribution of residuals of species intercept models of sensitivity and their relationship with scaled DBH, TWI and CII and deciduousness. Grey lines are linear model fits for statistically significant relationships at p<0.05. Species intercept model results are reported in main text.](display/residuals.png)

@@

![**Figure S14:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)

@@

![**Figure S15:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/cond_dep_dbh_twi.png)

![**Figure S16:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/cond_dep_cii_twi.png)

@@

![**Figure S17:** Correlation of modelled species sensitivities with Topographic Position Index](display/FigSI_TPI_preds.png)

@@

![**Figure S18:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TPI.](display/Fig4_tpi_3yrs.png)

@@

![**Figure S19:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)

@@

![**Figure S20:** Model results from three independent causal inquiry models of sensitivity following backdoor criteria for total causal effect of DBH, CII and TWI with species random effect on intercept and slope.](display/causal_inquiry.png)
